---
title: "ca_race"
output: html_document
---
<!-- Scatter plot of x = experience, y = income/wages, color = gender, shape = OCC group
Graph wages and non-wage income (PINCP-WAGP) too see if most inequality comes from non-wage income -->

```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo=FALSE, include = TRUE, warning = FALSE,
                      fig.align = "center", message = FALSE)

SF.xlim <- c(-123.1, -121.5)
SF.ylim <- c(37, 38.2)
LA.xlim <- c(-119, -116.7)
LA.ylim <- c(33, 34.5)

library(tidyverse)
library(plyr)
library(data.table)
library(knitr)
library(bit64)
library(scales)
library(maptools)
library(mapproj)
library(stringr)


new_kable <- function(...){
    kable(..., format.args = list(big.mark = ","))
}

# Read the raw data for this section

# filter.data <- fread("C:/Users/Owner/Documents/R/PUMS/filterData.csv")
map_data <- fread("C:/Users/Owner/Documents/R/PUMS/map_data.csv")


# Set up the maps
path <- "C:/Users/Owner/Documents/R/ACS/PUMAS/"

pumas <- readShapePoly(paste0(path,"cb_2015_06_puma10_500k.shp"), 
                       IDvar = "PUMACE10")

pumas.points <- data.table(fortify(pumas))

theme.simeon <- theme_gray() +
    theme(axis.title = element_text(family = "serif",
                                    size = 14,
                                    color = "blue4"),
          plot.background = element_rect(fill = "cornsilk",
                                         color = "black"),
          plot.title = element_text(family = "serif",
                                    face = "bold",
                                    size = 16,
                                    color = "blue4",
                                    hjust = 0.5),
          plot.subtitle = element_text(family = "serif",
                                    face = "bold",
                                    size = 10,
                                    color = "blue4",
                                    hjust = 0.5),
          plot.caption = element_text(family = "serif",
                                    face = "bold",
                                    size = 10,
                                    color = "blue4"),
          legend.background = element_rect(fill = "cornsilk",
                                           color = "blue4"),
          legend.title.align = 0.5,
          legend.title = element_text(family = "serif",
                                      size = 12,
                                      color = "blue4",
                                      face = "italic"),
          legend.key.size = unit(1.2, "lines"),
          legend.key.height = unit(0.8, "cm"),
          panel.background = element_rect(fill = "grey92",
                                          color = "blue4"))

theme_set(theme.simeon)

#Create label vector for races
race_labels = c("White" = "White",
                "Other" = "Other",
                "Asian" = "Asian",
                "African American" = "African \nAmerican")

edu_labels <- c("Post-Graduate" = "Post\nGraduate", 
                "Bachelors" = "Bachelors", 
                "Associates" = "Associates", 
                "HS Diploma" = "HS\nDiploma", 
                "HS-no Diploma" = "HS-no\nDiploma", 
                "No HS" = "No HS", 
                "None" = "None")
```


```{r Race data setup, results="hide"}
Race.data <- filter.data[, .(year, PUMA10, age, citizen, PINCP, PWGTP,
                             empStat, county, AGEP, WKHP,
                             race, Hispanic, sex, Occ)]
Race.data[, year := as.Date(year, "%Y-%m-%d")]
Race.data[, nage := round_any(AGEP, 7)]
Race.data[, white := ifelse(race == "White", "White", "Non-white")]
```

Let's start by getting a sense of the racial makeup of California. Broad categories are used for assigning Race in this analysis to both make the analysis simpler as well as increase sample sizes throughout. Our race category will be made up of: White, African American, Asian, and Other. The first three are explicit answers to the ACS question of Race, while the Other category includes all responses, (including some Hispanic), not in one of White, African American, or Asian. There is a large overlap between Hispanic and White as there is no specific Hispanic category. For this analysis I do not limit White to include only those respondents who do not indicate Hispanic ethnicity as well, and so leave responses as found in the raw ACS data.This means that the White category includes any respondent who considers their race to be White even if other categories, such as "origin" indicate Hispanic or other origin.

```{r "Racial makep of CA Adults"}

Race.data[, .(count = sum(PWGTP)), .(AGEP, race)][, .(percentage = count/sum(count), race), .(AGEP)] %>% 
    ggplot(aes(x = AGEP, y = percentage, fill = race)) +
    geom_area(alpha = 0.8, color = "black") +
    ggtitle("Racial Makeup of Adults in California",
            "Sample limited to observations between the ages of 18-75") +
    labs(fill = "Race",
         caption = "Jagged edges due to sampling sizes\nSimeon Markind") +
    scale_x_continuous("Age", breaks = seq(20,75,5), 
                       labels = seq(20,75,5)) +
    scale_y_continuous("Share of Population", labels = percent)
```


A short analysis of the Other category shows the makeup for our age buckets. Remember that the ACS RAC1P category does not include Hispanic as an option.

```{r}
other_data <- filter.data[race == "Other", .(count = sum(PWGTP)), .(RAC1P, age)][, .(RAC1P, percentage = count/sum(count)), .(age)]

other_data %>% ggplot(aes(x = age, y = percentage, fill = factor(RAC1P))) +
    geom_bar(stat = "identity",
             color = "black", size = 0.7, alpha = 0.8) +
    ggtitle("Racial Makeup of Other Category: Average values 2010-2014",
            "81% of Other category indicated Hispanic origin") +
    labs(caption = "Note: \"Tribe Specified\" refers to: \"American Indian and Alaska Native \ntribes specified; or American Indian or Alaska Native, not specified and no other races\"\nSimeon Markind",
         x = "Age") +
    scale_y_continuous("Share of Category", labels = percent) +
    scale_fill_discrete("Race", 
                        labels = c("3" = "American\nIndian only",
                                   "4" = "Alaska\nNative only",
                                   "5" = "Tribe\nSpecified",
                                "7" = "Hawaiian or\nPacific Islander",
                                   "8" = "Other",
                                "9" = "Two or more")) 
                                   
```

We can however look at the Hispanic percentage of the population using the HISP variable. In addition to making up the vast majority of the Other category Other subcategory, 33.5% of respondents who selected a race of White also indicated Hispanic heritage.

```{r "Hispanic percentage of Other", include = F, eval = F}
filter.data[!(RAC1P %in% c(1,2,6)), .(count = sum(PWGTP)), .(Hispanic)][, .(count/sum(count), Hispanic)]

filter.data[race == "White", .(count = sum(PWGTP)), .(Hispanic)][, .(count/sum(count), Hispanic)]
```


```{r "Hispanic Percentage Population"}
#Three year age buckets
Race.data[, .(PWGTP, Hispanic, age_buck = round_any(AGEP, 3))][, .(count = sum(PWGTP)), .(Hispanic, age_buck)][, .(Hispanic, percentage = count/sum(count)), .(age_buck)][Hispanic == "Hispanic"] %>% 
    ggplot(aes(x = age_buck, y = percentage)) +
    geom_line() + 
    scale_y_continuous("", label = percent, limits = c(0,0.5)) +
    ggtitle("Hispanic percentage of California population") +
    labs(caption = "2010-2014 average - Simeon Markind") +
    scale_x_continuous("Age", breaks = seq(20,75,5))
```


Before we dive into the data let's take a look at the mean incomes for each race over the period in constant 2014 USD for each race over the period. For our average income data we will simply look at respondents who are either self-employed or employed.

```{r "Mean income by Race and Age"}
table.data <- Race.data[empStat != "Unemployed",
                        .(mean_inc = round(weighted.mean(PINCP, PWGTP),2)), .(nage, race, sex)]

#dcast.data.table(table.data, sex + age ~ race, value.var = "mean_inc") # %>% new_kable(col.names = c("Gender", "Age Range", "African American",
#                          "Asian", "Other", "White"))
# Pretty big table, would be better as a graph of 8 lines
ymax <- max(table.data$mean_inc)

ggplot(table.data, aes(x = nage, y = mean_inc, color = race,
                       linetype = sex)) +
    geom_line(size = 1.2) +
    ggtitle(label = "Mean Income for California 2010-2014",
            subtitle = "Employed observations only, age below 70") +
    scale_x_continuous(name = "Age", limits = c(20, 70),
                       breaks = seq(20, 75, 5)) +
    scale_y_continuous(name = "2014 USD", labels = dollar,
                       limits = c(0, 1.1*ymax)) +
    scale_color_discrete(name = "Race") +
    scale_linetype_discrete("Gender") + 
    labs(caption = "Simeon Markind")
```

So the first thing you might ask is: "What on Earth is going on with Asian men and women?" (I have no idea - look into this?) The second thing you should note is the large income gap in later years, White men have larger incomes. Interestingly, White women also have the highest mean incomes in later years of all women, but the gap is not nearly as pronounced. Let's see if we can explain this gap a bit more as we dig deeper into the data.

Here we show a density plot of income by Race. What this plot shows is the percentage of each grouping, in our case Race, with an average income equal to our value on the x axis.

```{r "Income density by Race"}
## Find the weighted mean income for each education bin
bin_inc <- Race.data[, .(inc = weighted.mean(PINCP, PWGTP)),
                     .(race)]

Race.data[PINCP > 1000 & WKHP >= 35, .(PINCP, weight = PWGTP/sum(PWGTP)), .(race)] %>% 
    ggplot(aes(x = PINCP, group = race, fill = race)) +
    geom_density(aes(weight = weight), alpha = 0.5, 
                 size = 0.6) +
    scale_x_log10("Income", breaks = 10^c(3, 3.544068, 4, 4.544068,
                                          5, 5.544068, 6),
                  labels = dollar) +
    geom_vline(data = bin_inc,
               aes(xintercept = inc, color = race), size = 1) +
    ggtitle("Distribution of Income for CA by Race",
            "Vertical Lines show the mean income levels") +
    labs(caption = "Limited to observations working over 35 hours per week and income above $1,000\nSimeon Markind") +
    scale_y_continuous("Density") +
    scale_fill_hue("Race", labels = edu_labels) +
    scale_color_hue("Race", labels = edu_labels)
```


Let's now take a look at a map looking at average income for each race in California in 2014 so that we can get a sense of differentiation across the state.

```{r "Map of CA Income - Data"}
plot_data <- Race.data[year == as.Date("2014-01-01"), .(mean_inc = weighted.mean(PINCP, PWGTP)/1000, count = sum(PWGTP)), .(race, PUMA10)][count > 50][pumas.points, on = c("PUMA10" = "id"), allow.cartesian = T] 
```

```{r "Mapping CA Income by Race", fig.height=6, fig.width=6}
plot_data %>%     
ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = round(mean_inc,1)), color = "black") +
    facet_wrap(~ race) +
    ggtitle("Mean Income in California in 2014",
            "At least 50 observations per PUMA per Race") +
    scale_fill_distiller("Mean \nIncome", palette = "PiYG",
                         limits = c(0, 80)) +
    labs(x = "Longitude", y = "Latitude", caption = "Simeon Markind")
```


```{r "Employment Types be Race", include = F}
table.data <- Race.data[, .(count = sum(PWGTP)),
                        by = .(race, empStat)][, 
                    .(empStat, 
                      percentage = round(count/sum(count),2)*100),                          .(race)]

emp_type_bars <- table.data %>% 
    ggplot(aes(x = race, y = percentage, fill = empStat)) +
    geom_bar(stat = "identity", position = "stack", color = "black") +
    theme(legend.key.size = unit(1.5, "lines"))

## Not a very useful chart
```



```{r}
## Looking at white's vs all others for employed persons only

wnw.data <- Race.data[empStat != "Unemployed", 
                      .(Income = weighted.mean(PINCP, PWGTP, na.rm = T)),
                      by = .(year, white, citizen)]
```

Before we take a deeper look at the racial data, let's take a brief look at another factor which might be influencing income, citizenship status. Non-citizens may find it much more difficult to find employment than citizens, either naturalized or by birth.

```{r wnw plot}
ymax <- max(wnw.data$Income)

cit.plot <- ggplot(wnw.data, 
                   aes(x = year, y = Income, color = citizen,
                       linetype = white)) +
    geom_line(size = 1.3) +
    ggtitle(label = "Income for California 2010-2014",
            subtitle = "Employed observations only") +
    scale_x_date(name = "Year") +
    scale_y_continuous(name = "2014 USD", labels = dollar,
                       limits = c(0, 1.1*ymax)) +
    scale_color_discrete(name = "Citizenship") +
    scale_linetype_discrete("Race") + 
    labs(caption = "Simeon Markind")

cit.plot

```

Here we already see that by restricting the observations to employed only we find that non-white naturalized citizens make almost as much as white citizens born here. (non-white naturalized citizens actually make more than white naturalized citizens). What is interesting is that non-whites born here make substantially less than naturilized non-whites or born/naturalized whites. Non-citizens, unsurprisingly make the least. Now that we see a difference in the income levels of citizens vs. non-citizens, let's take a look at the makeup of our different racial classes in terms of citizenship status.

```{r "Percent Citizen by Race", fig.width=10}
cit_race_bars <- Race.data[, .(count = sum(PWGTP)), .(race, citizen, age)][, .(citizen, percentage = round(count/sum(count),4)*100), .(race, age)] %>% 
    ggplot(aes(x = age, y = percentage, fill = citizen)) +
    geom_bar(stat = "identity", alpha = 0.8, 
             color = "black", size = 0.6) +
    facet_wrap(~race, nrow = 2, ncol = 2) +
    theme(legend.position = "bottom") +
    ggtitle("The presence of such a large naturalized Asian population may explain the steep drop in Income",
            "Simeon Markind") +
    labs(x = "Age Bracket", y = "Percent", fill = "Citizenship Status")

cit_race_bars
```

As mentioned on the chart, the large proportion of the Asian population that is naturalized may play a part in the steep income declines of that community after age 40. It is likely that this population came to the US either later in life, (perhaps with adult children and therefore did not need to find high income jobs), or are less educated than younger Asian immigrants and therefore were only able to obtain lower paying employment than more recent immigrants. Below is a chart showing the unemployment rate for each race by age.

```{r "Unemployment Rate by Race"}
Race.data[, .(count = sum(PWGTP)), .(race, nage, empStat)][,
          .(empStat, percentage = round(count/sum(count),4)),
          .(race, nage)][nage %in% c(21:63) & empStat == "Unemployed",]%>% 
    ggplot(aes(x = nage, y = percentage, color = race)) +
    geom_line(size = 1.1) + 
    ggtitle("Average Percentage not employed 2010-2014",
            "Simeon Markind") + 
    scale_y_continuous(limits = c(0, 0.4), labels = percent) +
    labs(x = "Age", y = "", caption = "Note: May differ from official unemployment statistics") +
    scale_color_discrete("Race", 
                         labels = c("White" = "White",
                                    "Other" = "Other",
                            "Asian" = "Asian",
                            "African American" = "African \nAmerican"))
```


Just as with our gender data, with our race data there will also be differentiation in how people sort themselves into occupations.

```{r "Occupation binning", fig.width= 8, fig.height=5}
fields.data <- Race.data[!is.na(Occ), 
                           .(count = sum(PWGTP)),
                           .(race, Occ)][order(Occ, race)]

share.data <- fields.data[, .(Occ, share = round(count/sum(count),4)), by = .(race)]

occ.share.bars <- ggplot(share.data, aes(x = race, y = share, fill = Occ)) +
    geom_bar(stat = "identity", position = "stack", color = "black") + 
    coord_flip() +
    theme(legend.position = "bottom") +
    labs(title = "Share of Population by Occupation Type",
         subtitle = "Simeon Markind",
         y = NULL,
         fill = NULL) +
    scale_x_discrete("Race", labels = c("White" = "White",
                                        "Other" = "Other",
                                        "Asian" = "Asian",
                                        "African American" = "African \nAmerican")) +
    guides(fill = guide_legend(nrow = 6, reverse = T))

occ.share.bars

```

When looking at shares of population within each occupation bucket we see that the bars for Asian and White are similar while the bars for African American and Other are also similar in their distribution. Below we limit our sample further, looking at people who work in a select subgrouping of industries and are considered to be working full-time, in this case at least 35 hours per week.

```{r "Racial Makeup and mean Salaray of industries", fig.height=8, fig.width=10}
Race.data[!is.na(Occ), .(count = sum(PWGTP), inc = weighted.mean(PINCP, PWGTP)),
          .(race, Occ)][, .(percentage = count/sum(count), inc, race), .(Occ)] %>% 
    ggplot(aes(x = Occ, y = percentage, fill = factor(race,
                                                      levels = c("Asian", "Other", "White", "African American")))) +
    geom_bar(stat = "identity", color = "black", alpha = 0.8) +
    geom_label(aes(label = dollar(round(inc))),
               position = position_stack(vjust = 0.5), size = 4) +
    coord_flip() +
    ggtitle("Racial Share of Occupation Buckets",
            "Mean income for each Race shown as text") +
    labs(x = "Occupation", caption = "Simeon Markind") +
    scale_y_continuous("Percent", label = percent) +
    scale_fill_discrete("Race", labels = race_labels)
                        
                        
                            
```


```{r Differential income of select occupations, fig.width=8}
control.data <- Race.data[!is.na(Occ) &!is.na(nage) & WKHP >= 35 & nage <= 65,
                            .(count = sum(PWGTP), 
                              income = round(weighted.mean(PINCP,
                                                            PWGTP, 
                                                           na.rm = T),
                                             2)/1000),
                            .(race, Occ, nage)][order(Occ)]

control.data[Occ %in% c("Admin Support", "Legal", "Education", 
                        "Sales", "Healthcare", "Production")] %>% 
    ggplot(aes(x = nage, y = income, colour = race)) +
    geom_line(size = 1) +
    facet_wrap( ~ Occ, nrow = 2, ncol = 3) + 
    scale_x_continuous("Age") +
    scale_color_hue(name = "Race", 
                    labels = c("White" = "White", "Asian" = "Asian",
                               "Other" = "Other", "African American" =
                                   "African \nAmerican")) +
    scale_y_continuous("USD Thousands", labels = dollar) +
    ggtitle(label = "Incomes for Selected Occupation types",
            subtitle = "California 2010-2014 average") +
    labs(caption = "Simeon Markind")

```

Notice that income for Asians seems to follow the trend we saw earlier of declining after age 40 across all 6 industries seen here, but is most pronounced in the legal profession. A further analysis of Asian income will be included below. The income gap appears less pronounced in occupations with lower overall income, which is unsurprising. However, despite the income gap being less pronounced in terms of absolute dollars, as the next chart shows, in relative terms the income gap is highly prevalent and is higher at older ages across all races and industries. Of the data shown here, only Asians who work in Education do not see a constant increase in terms of relative income gap. 

<!-- Let's take a look at the income gap for all races and occupations
continuing to only look at people who work full time-->

```{r "Income Ratio by Race", fig.height=6}
# There must be a way to do this with long data but I don't know it
control.data %>% 
    dcast(Occ + nage ~ race, value.var = "income") %>% 
    mutate(ratio_AA = `African American`/White,
           ratio_Asian = Asian/White,
           ratio_Other = Other/White) %>% 
    select(Occ, nage, ratio_AA, ratio_Asian, ratio_Other) %>%
    filter(Occ %in% c("Admin Support", "Legal", "Education", 
                        "Sales", "Healthcare", "Production")) %>% 
    melt(id.vars = c("Occ", "nage")) %>% 
    ggplot(aes(x = nage, y = value, color = variable)) +
    geom_line() +
    facet_wrap(~ Occ, nrow = 2, ncol = 3) +
    ggtitle(label = "Income as a percentage of White's Income",
            subtitle = "California 2010-2014 Average full-time Employees: Simeon Markind") +
    labs(caption = "As age increases all but Asians in Education see \ndecreasing income relative to Whites",
         x = "Age",
         y = "Fraction of Income") +
    scale_color_hue("Race",
                    labels = c("ratio_AA" = "African \nAmerican",
                               "ratio_Asian" = "Asian",
                               "ratio_Other" = "Other"))
    
```

But looking at average income does not tell the whole story, we should also look at changes in average income across different Occupations for the same reasons we analyzed this data for each gender.

```{r "Income Growth by Race", fig.height=6, fig.width=7}
ratio.data <- Race.data[WKHP > 35 & !is.na(Occ) & nage <= 63,
                .(count = sum(PWGTP),
                  income = round(weighted.mean(PINCP, PWGTP, 
                                               na.rm = T),
                                             2)/1000),
                            .(year, race, Occ, nage)][order(year,nage)][count > 200, #Need at least 500 weighted entries to be included, otherwise the sample is too small
            .(year, growth_rate = income/lag(income)),
            .(race, Occ, nage)][, .(avg_growth = mean(growth_rate, na.rm = T)), .(race, Occ, nage)]
# Now let's plot it
inds <- c("Admin Support", "Legal", "Education", 
                        "Sales", "Healthcare", "Production")

ratio.data[!is.na(avg_growth) & avg_growth < 1.2 & Occ %in% inds] %>% 
# Remove older lawyers with higher growth rates who make the axes 
    # too wide to see variation for other years/professions
    ggplot(aes(x = nage, color = race, y = avg_growth)) +
    geom_line() +
    facet_wrap(~ Occ, nrow = 2, ncol = 3) +
    ggtitle(label = "On Average Growth rates of White people \nwere smaller for select Industries",
            subtitle = "Average Growth Rate for years 2010-2014, age bins 21-63") +
    labs(caption = "At least 200 weighted observations per point\nSimeon Markind",
         x = "Age",
         y = "Average Annual Income Growth Rate") +
    scale_color_hue("Race",
                    labels = c("African American" = "African \nAmerican", "Asian" = "Asian", "Other" = "Other", "White" = "White"))
```

I removed the growth rates for lawyers above the age of 60, as the Asian growth rate for that category was 1.56 which pushed the y-axes so high as to lose the ability to see variation in the other professions. When calculating my age, race, profession buckets for each year, 95% had populations greater than 200. I eliminate all observations for less than 200 weighted individuals as a check against potential outliers. Here is the same chart when looking at both race and gender.

```{r "Growth Rates for Race and Gender", fig.height=6}
ratio.data <- Race.data[WKHP > 35 & !is.na(Occ) & nage <= 63,
                .(count = sum(PWGTP),
                  income = round(weighted.mean(PINCP, PWGTP, 
                                               na.rm = T),
                                             2)/1000),
                            .(year, race, Occ, nage, sex)][order(year,nage)][count > 200, #Need at least 500 weighted entries to be included, otherwise the sample is too small
            .(year, growth_rate = income/lag(income)),
            .(race, Occ, nage, sex)][, .(avg_growth = mean(growth_rate, na.rm = T)), .(race, Occ, nage, sex)]
# Now let's plot it
inds <- c("Admin Support", "Legal", "Education", 
                        "Sales", "Healthcare", "Production")

ratio.data[!is.na(avg_growth) & Occ  %in% inds &
               avg_growth < 1.2 & avg_growth > 0.8] %>% 
# Remove older lawyers with higher growth rates who make the axes 
    # too wide to see variation for other years/professions
    ggplot(aes(x = nage, color = race, y = avg_growth, 
               linetype = sex)) +
    geom_line() +
    facet_wrap(~ Occ, nrow = 2, ncol = 3) +
    ggtitle(label = "Average Growth rates for men and women\ndid not differ dramatically for select industries",
            subtitle = "Average Growth Rate for years 2010-2014") +
    labs(caption = "At least 200 weighted observations per point\nSimeon Markind",
         x = "Age",
         y = "Average Annual Income Growth Rate") +
    scale_color_hue("Race",
                    labels = c("African American" = "African \nAmerican", "Asian" = "Asian", "Other" = "Other", "White" = "White")) +
    scale_linetype_discrete("Gender")

```

Here I limit the data shown on the plot to be in the range of (0.8, 1.2) which causes multiple observations from the legal profession to be removed in the interest of being able to clearly show the rest of the data. In this case removals were mainly too high, above 1.2, to fit on the plot. However, it is also of note that 28 year old males with race "Other" in the Legal occupation also do not appear as their average annual income growth rate was too low at 0.56.


```{r "Largest and Smallest Average Growth Rates", include=F, eval=F}
# find the max average avg_growth for each race and the associated Occ
# Unfortunately this method ends up with industries with incomplete series (only one race, or missing certain age buckets)

lg_rates <- ratio.data[, .(growth = mean(avg_growth, na.rm = F)), .(Occ, race)][order(-growth)][!is.na(growth), head(.SD,1), .(race)][, Occ]

sm_rates <- ratio.data[, .(growth = mean(avg_growth, na.rm = F)), .(Occ, race)][order(-growth)][!is.na(growth), tail(.SD,1), .(race)][, Occ]
```


```{r "Table of average growth rates per Occ and Race"}
ratio.data[, .(growth = round(mean(avg_growth, na.rm = T),3)), .(race, Occ)] %>% 
    dcast(Occ ~ race, value.var = "growth") %>% 
    new_kable(caption = "Average income growth 2010-2014", digits = 3, col.names = c("Occupation", "African American", "Asian", "Other", "White"))
```


```{r Differential binning of occupations plot, include=F, eval=F}
raceBin.plot <- ggplot(raceBin.data, 
                       aes(x = Occ, y = income, shape = race)) +
    geom_jitter(aes(color = race)) +
    scale_x_discrete(name = "Occupation",
                     labels = c("Computer/Math" = "Technology",
                                "Admin Support" = "Admin",
                                "Farm/Fish/Forestry" = "Agriculture",
                                "Healthcare" = "Healthcare",
                                "Sales" = "Sales",
                                "Education" = "Education")) +
    scale_y_continuous(name = "Income", labels = dollar) +
    ggtitle(label = "Distribution of Income",
            subtitle = "Wighted Mean: Select Occupation Groups") +
    labs(caption = "Simeon Markind") +
    scale_color_discrete(name = "Race") +
    scale_shape_discrete(name = "Race")

raceBin.plot

## It appears that white people make more within the different occupation types of average but not outwardly more, except in Sales where there is a clear trend of white outearning non-whites. This does not mean that across all occupation types there is not a huge differentiation among races, but it does point to race not being the sole factor, or perhaps even the major factor in the income distribution.
```

```{r "Racial Migration in CA", results="hide"}
## Start by seeing who is actually moving in/around the state
mobility.data <- filter.data[, .(county, MIGSP05, MIGSP12, Occ, PUMA00, PUMA10,
                                 sex, age, race, year, PINCP, PWGTP)]
# for MIGS, NA means didn't move, 6 means moved within CA, -9 is NA
# a number greater than 6 is outside of CA. 
# a three digit number is outside of the USA

mobility.data[year <= as.Date("2011-01-01"),
              mobility := ifelse(is.na(MIGSP05) | MIGSP05 == -9, "Non-movers",
                                 ifelse(MIGSP05 == 6, "In State",
                                        ifelse(MIGSP05 >= 100, "Foreign", "In US")))]

mobility.data[year > as.Date("2011-01-01"),
              mobility := ifelse(is.na(MIGSP12) | MIGSP12 == -9, "Non-movers",
                                 ifelse(MIGSP12 == 6, "In State",
                                        ifelse(MIGSP12 >= 100, "Foreign", "In US")))]

mobility.data[, c("MIGSP05", "MIGSP12") := NULL]
```

```{r "Mobility Table"}
avg_movers <- mobility.data[, .(count = sum(PWGTP)), .(race, mobility)][, .(mobility, percent = round(count/sum(count),4)*100), .(race)]

avg_movers %>% 
    dcast(mobility ~ race, value.var = "percent") %>% 
    new_kable(caption = "Percentages", col.names = c("Mobility Status", "African American" ,"Asian", "Other", "White"))

```

```{r "Bros moving to SF", fig.width=9.5}
sf_counties <- c("Sonoma", "Napa", "Solano", "Marin", "San Francisco",
                 "San Mateo", "Santa Cruz", "Santa Clara", "Alameda",
                 "Contra Costa")

la_counties <- c("Los Angeles", "Orange", "Riverside", "San Bernardino")

county_level_mobility <- mobility.data[, .(count = sum(PWGTP),
                                           inc = weighted.mean(PINCP, PWGTP)), .(county, race, mobility)][,
        .(race, mobility, inc = weighted.mean(inc, count),
          percentage = round(count/sum(count),4)*100),
        .(county)]

in_state_movers <- county_level_mobility %>% 
    filter(county %in% sf_counties, mobility == "In State") %>% 
    select(race, county, percentage) %>% 
    ggplot(aes(x = county, y = percentage, fill = race)) +
    geom_bar(stat = "identity", color = "black") +
    ggtitle("In State movers to Bay area counties",
            "Average percentage of county residents who lived elsewhere in CA in the previous year") +
    labs(x = "County", y = "Percentage", 
         caption = "Simeon Markind") +
    scale_fill_discrete("Race", 
                   labels = c("African American" = "African \nAmerican",
                              "White" = "White",
                              "Other" = "Other",
                              "Asian" = "Asian"))

in_state_movers
```

```{r "In US Movers to SF", fig.width=9.5}
in_us_movers <- county_level_mobility %>% 
    filter(county %in% sf_counties, mobility == "In US") %>% 
    select(race, county, percentage) %>% 
    ggplot(aes(x = county, y = percentage, fill = race)) +
    geom_bar(stat = "identity", color = "black") +
    ggtitle("Domestic movers to Bay area counties",
            "Average percentage of county residents who lived in the US but not in CA the previous year") +
    labs(x = "County", y = "Percentage", 
         caption = "Simeon Markind") +
    scale_fill_discrete("Race", 
                   labels = c("African American" = "African \nAmerican",
                              "White" = "White",
                              "Other" = "Other",
                              "Asian" = "Asian"))
in_us_movers
```

As we can see from the bar charts showing domestic movers to the Bay Area the majority of people moving into the Bay Area are White with the vast majority either White or Asian. The average income for Whites and Asians moving to (or within) the Bay Area from within CA were \$55,400 and \$57,570. The average income for people who moved within the US from outside of CA were \$56,140 and \$59,550.  


```{r "SF Income by Race", fig.height=6}
plot_data %>%     
ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = round(mean_inc,1)), color = "black") +
    facet_wrap(~ race) +
    coord_map(xlim = SF.xlim, ylim = SF.ylim) +
    ggtitle("Mean Income in Bay Area in 2014 - PUMA",
            "At least 50 observations per PUMA per Race") +
    scale_fill_distiller("Mean \nIncome", palette = "PiYG",
                         limits = c(0, 120)) +
    labs(x = "Longitude", y = "Latitude", caption = "Simeon Markind")
```

```{r "LA Income by Race", fig.height=6}
plot_data %>%     
ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = round(mean_inc,1)), color = "black") +
    facet_wrap(~ race) +
    coord_map(xlim = LA.xlim, ylim = LA.ylim) +
    ggtitle("Mean Income in the LA Area in 2014",
            "At least 50 observations per PUMA per Race") +
    scale_fill_distiller("Mean \nIncome", palette = "PiYG",
                         limits = c(0, 120)) +
    labs(x = "Longitude", y = "Latitude", caption = "Simeon Markind")
```


```{r Heat map of population CA percentage white in 2014}
raceMap.data <- map_data[year == as.Date("2014-01-01"),
                          .(count = sum(PWGTP)), .(PUMA10, white)]

raceMap.data <- raceMap.data[, .(white, percent = (sum(white == "White")*count)/sum(count)),
                     by = .(PUMA10)][white == "White",]

raceMap.data <- merge(raceMap.data, pumas.points, 
               by.x = "PUMA10", by.y = "id", all = T)

raceMap.plot <- ggplot(raceMap.data, 
                       aes(x = long, y = lat, group = group)) +
    geom_polygon(color = "black", aes(fill = percent*100), alpha = 1.0) +
    scale_fill_distiller(name = "Percentage", palette = "BrBG", 
                         limits = c(0,95), position = "bottom") +
    coord_map() +
    ggtitle("PUMA Level percentage white - 2014", 
            subtitle = "Simeon Markind") +
    scale_x_continuous(name = "Longitutde") +
    scale_y_continuous(name = "Latitude")

raceMap.plot
```

Unsurprisingly California, especially rural California is majority white. Each shaded region on the plot is equal to one Public Use Microdata Area, a region of at least 100,000 people, so we can see that inland California is quite rural. The presence of large uninhabited park lands such as Yosemite, Sequioa, and Death Valley can clearly be felt in the eastern part of the state.

```{r SF PUMAs white}
sfWhite.plot <- ggplot() + geom_polygon(data = raceMap.data, 
                 mapping = aes(x = long, y = lat, group = group, 
                               fill = percent*100),
                 color = "black") +
 #   geom_point(data = ca.cities, 
 #                mapping = aes(x = long, y = lat, size = pop),
 #                 color = "coral1") +
 #   scale_size_continuous(name = "Population") +
 #   geom_label(data = ca.cities, nudge_x = 0.01, nudge_y = 0.01, 
 #              mapping = aes(x = long, y = lat, label = name),
 #              check_overlap = TRUE) +
    scale_fill_distiller(name = "Percentage", palette = "BrBG", 
                         limits = c(0,95), position = "bottom") +
    coord_map(xlim = SF.xlim, ylim = SF.ylim) +
    ggtitle("PUMA Level percentage white - 2014", 
            subtitle = "SF Bay area") +
    scale_x_continuous(name = "Longitutde") +
    scale_y_continuous(name = "Latitude")

sfWhite.plot
```

```{r "SF White Male 18-35"}
white_youth_male_frac <- map_data[year == as.Date("2014-01-01"), .(count = sum(PWGTP)), .(sex, AGEP, PUMA10, white)][, .(frac = sum((sex == "Male" & white == "White" & AGEP %in% c(18:35))*count)/sum(count)), .(PUMA10)][pumas.points, on = c("PUMA10" = "id")] #join with the mapping data so we can make the plots

# map_data[year == as.Date("2014-01-01"), .(count = sum(PWGTP)), .(sex, AGEP, white)][, .(frac = sum((sex == "Male" & white == "White" & AGEP %in% c(18:35))*count)/sum(count))]

bay_wymf <- white_youth_male_frac %>% 
    ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = frac*100), alpha = 0.9, color = "black") + 
    coord_map(xlim = SF.xlim, ylim = SF.ylim) +
    scale_fill_distiller("Percentage", palette = "OrRd", direction = 1,
                         limits = c(0, 15)) +
    scale_x_continuous("Longitutde") +
    scale_y_continuous("Latitude") +
    ggtitle(label = "San Francisco has a lot of Bros",
            subtitle = "Percent of population made up by white males 18-35 as of 2014") +
    labs(caption = "CA average is 7.9% \nSimeon Markind")

bay_wymf

# sf_wymf <- white_youth_male_frac %>% 
#     ggplot(aes(x = long, y = lat, group = group)) +
#     geom_polygon(aes(fill = frac*100), alhpa = 0.9, color = "black") + 
#     coord_map(xlim = c(-122.6, -122.2), ylim = c(37.6, 37.9)) +
#     scale_fill_distiller("Percentage", palette = "OrRd", direction = 1,
#                          limits = c(0, 20)) +
#     scale_x_continuous("Longitutde") +
#     scale_y_continuous("Latitude") +
#     ggtitle(label = "San Francisco has a lot of Bros",
#             subtitle = "Percent of population made up by white males 18-35 as of 2014") +
#     labs(caption = "CA average is 11% \nSimeon Markind")
# 
# sf_wymf
                                                                
```

```{r "SF Asian frac"}
map_data[year == as.Date("2014-01-01"), .(count = sum(PWGTP)), .(race, PUMA10)][, .(race, frac = round(count/sum(count),3)), .(PUMA10)][race == "Asian"][pumas.points, on = c("PUMA10" = "id")] %>% 
    ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = frac*100), alpha = 0.9, color = "black") + 
    coord_map(xlim = SF.xlim, ylim = SF.ylim) +
    scale_fill_distiller("Percentage", palette = "OrRd", 
                         direction = 1) +
    scale_x_continuous("Longitutde") +
    scale_y_continuous("Latitude") +
    ggtitle(label = "San Francisco percentage Asian") +
    labs(caption = "CA average is 14% \nSimeon Markind")
```



```{r LA PUMAS White}
LAWhite.plot <- ggplot(raceMap.data, 
                       aes(x = long, y = lat, group = group)) +
    geom_polygon(color = "black", aes(fill = percent*100), alpha = 1.0) +
    scale_fill_distiller(name = "Percentage", palette = "BrBG", 
                         limits = c(0,95), position = "bottom") +
    coord_map(xlim = LA.xlim, ylim = LA.ylim) +
    ggtitle("PUMA Level percentage white - 2014", 
            subtitle = "LA area") +
    scale_x_continuous(name = "Longitutde") +
    scale_y_continuous(name = "Latitude")

LAWhite.plot
```

```{r LA Asian frac}
map_data[year == as.Date("2014-01-01"), .(count = sum(PWGTP)), .(race, PUMA10)][, .(race, frac = round(count/sum(count),3)), .(PUMA10)][race == "Asian"][pumas.points, on = c("PUMA10" = "id")] %>% 
    ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(aes(fill = frac*100), alpha = 0.9, color = "black") + 
    coord_map(xlim = LA.xlim, ylim = LA.ylim) +
    scale_fill_distiller("Percentage", palette = "OrRd", 
                         direction = 1) +
    scale_x_continuous("Longitutde") +
    scale_y_continuous("Latitude") +
    ggtitle(label = "Los Angeles percentage Asian") +
    labs(caption = "CA average is 14% \nSimeon Markind")
```

